Testing gene methylation enrichment analysis approaches using simulated data

### 2023-10-16 figs/GSE158422_simulate.Rmd
bugfix

  # divide gene sets between hyper and hypomethylated
  nset <- floor(length(genesets)/2)
  set.seed(seed) ; gtup <-sample(genesets,nset)
  set.seed(seed) ; gtdn <- sample(setdiff(genesets,gtup),nset)
  gup <- unname(unlist(gtup))
  gdn <- unname(unlist(gtdn))
  # make probe-gene vector
  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
  names(probe2gene) <- rownames(myann)
  probe2gene <- unlist(probe2gene)
  # select probes hypermethylated
  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
  pup <- names(probe2gene[which(probe2gene %in% gup2)])
  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
  # select probes hypomethylated
  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
  pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pup)*frac_probes))
  # divide samples between ctrl and case
  ncols <- ncol(mval)
  maxgroupsize=floor(ncols/2)
  if ( groupsize > maxgroupsize ) { stop("groupsize cannot be larger than half the ncols of mval") }
  set.seed(seed) ; ctrl <- sample(1:ncols,groupsize)
  set.seed(seed) ; case <- sample(setdiff(1:ncols,ctrl),groupsize)
  mval_ctrl <- mval[,ctrl]
  mval_case <- mval[,case]
  # incorporate altered signals - change by +1 or -1
  mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] + delta
  mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] - delta
  mval2 <- cbind(mval_ctrl,mval_case)
  result <- list("mval"=mval2,"probes up"=pup2,"probes down"=pdn2,
    "genes up"=gup2,"genes down"=gdn2,
    "genesets up"=gtup,"genesets down"=gtdn)
  return(result)
}

@@

simgsa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
  pup3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC>0))
  pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
  # convert gene sets to entrez
  suppressWarnings(suppressMessages({ gene2entrez <- mapIds(org.Hs.eg.db, gene_catalog, 'ENTREZID', 'SYMBOL') }))
  gsets_entrez <- lapply(gsets,function(gs) {
    gs2 <- unique(gene2entrez[names(gene2entrez) %in% gs])
    gs2 <- gs2[!is.na(gs2)]
    return(gs2)
  })
  suppressWarnings(suppressMessages({
    gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez)
    gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez)
  }))
  gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
  gsig_dn3 <- rownames(subset(gsadn3,FDR<0.05))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(gsadn3)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

groupsize=5
delta=1
num_dm_sets=50

@@

gres <- lapply(1:6,function(i) {
  simgsa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
})

@@

lares <- mclapply(1:6,function(i) {
  simla(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
},mc.cores=6)

@@

alres <- mclapply(1:6,function(i) {
  simal(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
},mc.cores=6)

@@

aares <- mclapply(1:6,function(i) {
  simaa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
},mc.cores=6)

@@

almres <- mclapply(1:6,function(i) {
  simalm(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
},mc.cores=6)

>>>>

  # divide gene sets between hyper and hypomethylated
  nset <- floor(length(genesets)/2)
  set.seed(seed) ; gtup <-sample(genesets,nset)
  set.seed(seed) ; gtdn <- sample(setdiff(genesets,gtup),nset)
  gup <- unname(unlist(gtup))
  gdn <- unname(unlist(gtdn))
  # make probe-gene vector
  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
  names(probe2gene) <- rownames(myann)
  probe2gene <- unlist(probe2gene)
  # select probes hypermethylated
  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
  pup <- names(probe2gene[which(probe2gene %in% gup2)])
  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
  # select probes hypomethylated
  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
  pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pdn)*frac_probes))
  # divide samples between ctrl and case
  ncols <- ncol(mval)
  maxgroupsize=floor(ncols/2)
  if ( groupsize > maxgroupsize ) { stop("groupsize cannot be larger than half the ncols of mval") }
  set.seed(seed) ; ctrl <- sample(1:ncols,groupsize)
  set.seed(seed) ; case <- sample(setdiff(1:ncols,ctrl),groupsize)
  mval_ctrl <- mval[,ctrl]
  mval_case <- mval[,case]
  # incorporate altered signals - change by +1 or -1
  mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] + delta
  mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] - delta
  mval2 <- cbind(mval_ctrl,mval_case)
  result <- list("mval"=mval2,"probes up"=pup2,"probes down"=pdn2,
    "genes up"=gup2,"genes down"=gdn2,
    "genesets up"=gtup,"genesets down"=gtdn)
  return(result)
}

@@

simgsa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
  pup3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC>0))
  pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
  if ( length(pup3) < 250 ) { pup3 <- head(rownames(subset(dm3, logFC > 0)), 250) }
  if ( length(pdn3) < 250 ) { pdn3 <- head(rownames(subset(dm3, logFC < 0)), 250) }
  # convert gene sets to entrez
  suppressWarnings(suppressMessages({ gene2entrez <- mapIds(org.Hs.eg.db, gene_catalog, 'ENTREZID', 'SYMBOL') }))
  gsets_entrez <- lapply(gsets,function(gs) {
    gs2 <- unique(gene2entrez[names(gene2entrez) %in% gs])
    gs2 <- gs2[!is.na(gs2)]
    return(gs2)
  })
  suppressWarnings(suppressMessages({
    gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez)
    gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez)
  }))
  gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
  gsig_dn3 <- rownames(subset(gsadn3,FDR<0.05))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(gsadn3)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

groupsize=3
delta=0.6
num_dm_sets=50

@@

gres <- lapply(1:100,function(i) {
  simgsa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
})

@@

lares <- mclapply(1:100,function(i) {
  simla(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
},mc.cores=6)

@@

alres <- mclapply(1:100,function(i) {
  simal(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
},mc.cores=6)

@@

aares <- mclapply(1:100,function(i) {
  simaa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
},mc.cores=6)

@@

almres <- mclapply(1:100,function(i) {
  simalm(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
},mc.cores=6)
